### 2014-01-16 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
Minor edits to dpp and reply manuscripts.

The disparity between the results of the \empModelOld model and the new models
is even more pronounced when we look at the 10 divergence models
(\divTimeIndexVector) estimated to have the highest probability under each of the
models
(Figures~S\labelcref{figDivModelsDPP,figDivModelsDPPInform,figDivModelsDPPSimple,figDivModelsUniform,figDivModelsOld}).
Again, the new models estimate more divergences, a large amount of posterior
uncertainty, and an order of magnitude smaller probability for their respective
MAP model when compared to the \empModelOld model.

Figure \ref{figNegrosPanay} shows the estimated posterior probability
distribution over the number of divergence events when the data from the
nine-taxon pairs from the Islands of Negros and Panay are analyzed under the DPP
models that sample over unordered (\npModelDPP) and ordered
(\npModelDPPOrdered) models of divergence.
The result are similar under both models and, again, yield a large amount of
uncertainty about the number of divergence events that is similar to the prior
uncertainty.
Both models estimate a similar set of 10 unordered divergence models with the
highest posterior probability
(Figures~S\labelcref{figDivModelsNP,figDivModelsNPOrdered}).

@@

The modifications to the model also improve the method's power to detect random
variation in divergence times, reducing the tendency to estimate clustered
divergences
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).
Our results are similar to those of \citet{Oaks2012} in that we also find
the \msb model to infer strong support for clustered divergences when
divergences are random over quite broad timescales
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).
Our results expand on this by showing that this behavior is consistent
across a range of conditions underlying the data.
The new method \dppmsbayes has greater power to detect random temporal variation in
divergences, is less prone to spurious inference of clustered divergence
models, and much less likely to incorrectly infer such models with strong
support
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).

By evaluating a model intermediate between the old and new implementation
(\modelUshaped), we were able to determine the relative effects of our
modifications to the model.
Across all of our analyses, the results show that using better priors on
divergence-time and demographic parameters alone does improve the performance
of the method.
The magnitude of the bias toward inferring support for the one-divergence
model when there are model violations is reduced when the exponential
priors are used in place of the uniform priors
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
Furthermore, better priors on parameters improves the method's power to detect
temporally random divergences
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).
Throughout our analyses, the intermediate model (\modelUshaped) performs better
than the \msb model, but not as well as the models with alternative priors on
divergence models.
This suggests, as predicted by \citet{Oaks2012}, that the bias of \msb toward
models of temporally clustered divergences is caused by a combination of
(1) small marginal likelihoods of models with more \divTime{} parameters due to
uniform priors on divergence-time and demographic parameters and
(2) the U-shaped prior on divergence models giving low prior density to models
with intermediate numbers of divergence parameters.
The former essentially rules out models with many \divTime{} parameters, which
causes the latter to act like an "L-shaped" prior with a spike of
prior density on the one-divergence model.
Given the parameter richness of the model and the relatively small amount of
information in the summary statistics, it is not surprising that the
combination of these two factors can create a strong tendency to infer
clustered models of divergence.

While our modifications improve the behavior of the model, we urge caution when
using the method and interpreting its results.
The method attempts to approximate the posterior of a very parameter-rich model
using relatively little information from the data.
For example, when applied to the dataset of 22 taxon pairs from the Philippines
\citep{Oaks2012} the model can have as many as 604--625 free parameters
(depending on \divTimeNum), and samples over 1002 unordered divergence models.
Even under the simplest possible model allowed under our new implementation,
the model still has 471--492 free parameters.
Furthermore, the stochastic coalescent and mutational processes being modeled
predict a large amount of variation in possible datasets even when the
parameter values are known.
The richness and stochastic nature of the model makes for a difficult inference
problem, especially when using a small number of summary statistics calculated
from the sequence alignments of each taxon pair.
The population-genetic summary statistics used by our method contain little
information about many of the free parameters in the model.
Thus, we expect the improved method will still be sensitive to priors, and the
power, while improved, may still be low.
While we see much less prior sensitivity under the new model compared to those
observed by \citet{Oaks2012}, we still see an effect when comparing the results
of the empirical data analyzed under a diffuse (\empModelDPP) and informative
(\empModelDPPInform) divergence-time prior (Figure~\ref{figPhilippines} C
versus D).
The fact that the posterior shifts toward the prior under the informative prior
suggests that the shift away from the prior toward fewer divergence events
under the diffuse prior might still be caused by small marginal likelihoods
of models with more divergence-time parameters (Figure~\ref{figPhilippines}).

>>>>

The disparity between the results of the \empModelOld model and the new models
is even more pronounced when we look at the 10 divergence models
(\divTimeIndexVector) estimated to have the highest probability under each of the
models
(Figures~S\labelcref{figDivModelsDPP,figDivModelsDPPInform,figDivModelsDPPSimple,figDivModelsUniform,figDivModelsOld}).
Again, the new models estimate more divergences, a large amount of posterior
uncertainty, and an order of magnitude smaller probability for their respective
MAP-divergence model when compared to the \empModelOld model
(Figures~S\labelcref{figDivModelsDPP,figDivModelsDPPInform,figDivModelsDPPSimple,figDivModelsUniform,figDivModelsOld}).

Figure \ref{figNegrosPanay} shows the estimated posterior probability
distribution over the number of divergence events when the data from the
nine-taxon pairs from the Islands of Negros and Panay are analyzed under the DPP
models that sample over unordered (\npModelDPP) and ordered
(\npModelDPPOrdered) models of divergence.
The results are similar under both models and, again, yield a large amount of
uncertainty about the number of divergence events that is similar to the prior
uncertainty.
Both models estimate a similar set of 10 unordered divergence models with the
highest posterior probability
(Figures~S\labelcref{figDivModelsNP,figDivModelsNPOrdered}).

@@

Our modifications also improve the method's power to detect random
variation in divergence times, reducing the tendency to estimate clustered
divergences
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).
Our results are similar to those of \citet{Oaks2012} in that we find
\msb will often infer strong support for clustered divergences when
divergences are random over quite broad timescales
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).
Our results expand on this by showing that this behavior is consistent
across a range of conditions underlying the data.
The new method \dppmsbayes has greater power to detect random temporal variation in
divergences, is less prone to spurious inference of clustered divergence
models, and much less likely to incorrectly infer such models with strong
support
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).

By evaluating a model intermediate between the old and new implementation
(\modelUshaped), we were able to determine the relative effects of our
modifications to the model.
Across all of our analyses, the results show that using better priors on
divergence-time and demographic parameters alone does improve the performance
of the method.
The magnitude of the bias toward inferring support for the one-divergence
model when there are model violations is reduced when the exponential
priors are used in place of the uniform priors
(Figure~\ref{figValidationModelChoicePsi} \&
\ref{figValidationModelChoiceOmega}).
Furthermore, using exponential priors improves the method's power to detect
temporally random divergences
(Figures~\labelcref{figPowerPsiOld4,figPowerPsiProbOld4,figPowerOmegaOld4,figPowerOmegaProbOld4}).
Throughout our analyses, the intermediate model (\modelUshaped) performs better
than the \msb model, but not as well as the models with alternative priors on
divergence models.
This suggests, as predicted by \citet{Oaks2012}, that the tendency of \msb to
erroneously support models of temporally clustered divergences is caused by a
combination of
(1) small marginal likelihoods of models with more \divTime{} parameters due to
uniform priors on divergence-time and demographic parameters and
(2) the U-shaped prior on divergence models giving low prior density to models
with intermediate numbers of divergence parameters.
The former essentially rules out models with many \divTime{} parameters, which
causes the latter to act like an "L-shaped" prior with a spike of
prior density on the one-divergence model.
Given the parameter richness of the model and the relatively small amount of
information in the summary statistics, it is not surprising that the
combination of these two factors can create a strong tendency to infer
clustered models of divergence.

While our modifications improve the behavior of the model, we urge caution when
using the method and interpreting its results.
The method attempts to approximate the posterior of a very parameter-rich model
using relatively little information from the data.
For example, when applied to the dataset of 22 taxon pairs from the Philippines
\citep{Oaks2012} the model has as many as 604--625 free parameters
(depending on \divTimeNum), and samples over 1002 unordered divergence models.
Even under the simplest possible model allowed under our new implementation,
the model still has 471--492 free parameters.
Furthermore, the stochastic coalescent and mutational processes being modeled
predict a large amount of variation in possible datasets even when the
parameter values are known.
The richness and stochastic nature of the model makes for a difficult inference
problem, especially when using a small number of summary statistics calculated
from the sequence alignments of each taxon pair.
The population-genetic summary statistics used by our method contain little
information about many of the free parameters in the model.
Thus, we expect the improved method will still be sensitive to priors, and the
power, while improved, may still be low.
While we see much less prior sensitivity under the new model compared to those
observed by \citet{Oaks2012}, we still see an effect when comparing the results
of the empirical data analyzed under a diffuse (\empModelDPP) and informative
(\empModelDPPInform) divergence-time prior (Figure~\ref{figPhilippines} C
versus D).
The fact that the posterior shifts toward the prior under the informative prior
suggests that the shift away from the prior toward fewer divergence events
under the diffuse prior might still be caused by small marginal likelihoods
of models with more divergence-time parameters (Figure~\ref{figPhilippines}).

### 2014-01-16 manuscripts/reply/reply.tex
Updating reply ms.
Another working title.
Minor edits to dpp and reply manuscripts.

\newcommand{\msTitle}{Nuh uh: A reply to Hickerson et al.
\xspace}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

{\sffamily
    \noindent\textbf{ABSTRACT} \\
    \noindent Biogeographers frequently seek to explain population and species
    differentiation on geographical phenomena.
    Establishing that a set of population splitting events occurred
    at the same time can be a potentially persuasive argument that a set of taxa
    were affected by the same geographic events.
    Unfortunately, estimating divergence times precisely when one lacks
    precise information about the rate of molecular evolution is a notoriously
    difficult problem in evolutionary biology.
    \citep{Huang2011} introduced an approximate Bayesian approach which can be
    used to estimate the probabilities of models in which multiple sets of taxa
    diverge at the same time. 
    This method (implemented in the software \msb) does not require a common
    set of character data that links all of the taxa under investigation, but
    does require the relative mutation rates among the taxa and loci to be
    assumed \emph{a priori}.
    Recently, \citet{Oaks2012} used this model-choice framework to 
    study 22 pairs of vertebrate lineages which are distributed across
    the Philippines; they also studied the behavior of the \msb approach
    using computer simulations.
    \citet{Oaks2012} found that the model was very sensitive to the prior and
    that the inference method had low power to detect variation in divergences
    times.
    These results were not surprising, in light of a rich statistical literature
    showing that the marginal likelihood of a model is sensitive to the
    use of vague priors.
    Because this sensitivity to prior assumptions affects the crucial insights 
    that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
    that users of the approach should carefully assess the robustness of their 
    conclusions to different priors.
    According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
    was due to excessively broad priors on divergence times, leading to 
    inadequate numbers of simulation replicates.
    They proposed a new model-averaging approach that uses narrow, empirically
    informed uniform priors.
    Here we demonstrate that the approach of \citet{Hickerson2013} is
    dangerous in the sense that the empirically-derived priors often
    exclude from consideration the true values of the models' parameters.
    On a more fundamental level, we question the value of adopting an empirical
    Bayesian stance for this model-choice problem, because it can mislead model
    posterior probabilities, which are inherently measures of belief in the
    models after prior knowledge is updated by the data.
    The robust Bayesian approach of conducting analyses under a variety
    of priors can reveal prior sensitivity and communicate which assumptions
    underlie model inference.
    Furthermore, simulations provide insight into the temporal resolution of
    the method, which in turn helps guide interpretation of results.

>>>>

\newcommand{\msTitle}{Implications of uniformly distributed, empirically
    informed priors for phylogeographical model selection: A reply to Hickerson
    et al.
\xspace}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

{\sffamily
    \noindent\textbf{ABSTRACT} \\
    \noindent Biogeographers frequently seek to explain population and species
    differentiation on geographical phenomena.
    Establishing that a set of population splitting events occurred
    at the same time can be a potentially persuasive argument that a set of taxa
    were affected by the same geographic events.
    Unfortunately, estimating divergence times precisely when one lacks
    precise information about the rate of molecular evolution is a notoriously
    difficult problem in evolutionary biology.
    \citet{Huang2011} introduced an approximate Bayesian approach which can be
    used to estimate the probabilities of models in which multiple sets of taxa
    diverge at the same time. 
    This method (implemented in the software \msb) does not require a common
    set of character data that links all of the taxa under investigation, but
    does require the relative mutation rates among the taxa and loci to be
    assumed \emph{a priori}.
    Recently, \citet{Oaks2012} used this model-choice framework to 
    study 22 pairs of vertebrate lineages which are distributed across
    the Philippines; they also studied the behavior of the \msb approach
    using computer simulations.
    \citet{Oaks2012} found that the model was very sensitive to the prior and
    that the inference method had low power to detect variation in divergences
    times.
    These results were not surprising, in light of a rich statistical literature
    showing that the marginal likelihood of a model is sensitive to the
    use of vague priors.
    Because this sensitivity to prior assumptions affects the crucial insights 
    that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
    that users of the approach should carefully assess the robustness of their 
    conclusions to different priors.
    According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
    was due to excessively broad priors on divergence times, leading to 
    inadequate numbers of simulation replicates.
    They proposed a new model-averaging approach that uses narrow, empirically
    informed uniform priors.
    Here we demonstrate that the approach of \citet{Hickerson2013} is
    dangerous in the sense that the empirically-derived priors often
    exclude from consideration the true values of the models' parameters.
    On a more fundamental level, we question the value of adopting an empirical
    Bayesian stance for this model-choice problem, because it can mislead model
    posterior probabilities, which are inherently measures of belief in the
    models after prior knowledge is updated by the data.
    The robust Bayesian approach of conducting analyses under a variety
    of priors can reveal prior sensitivity and communicate which assumptions
    underlie model inference.
    Furthermore, simulations provide insight into the temporal resolution of
    the method, which in turn helps guide interpretation of results.
